---
title: "Measuring and Visualizing Retention"
subtitle: "Flow Tracking"
author: "Ann Nakamura"
date:  "2/4/2019"
output:
  github_document:
    toc: true
    toc_depth: 2
    number_sections: TRUE
---

<h1 style="color: #154c79">Using Network Diagrams</h1>

# Use Case

Create a quick visualization, using network diagrams and alluvial plots to examine growth and retention for group members who enter and exit at random points in time. Uses a [Fake cohort](https://raw.githubusercontent.com/AMNakamura/miscellanea/master/datasets/FakeCohort1.txt) for demonstration purposes. 


```{r setGrps, message=FALSE,warning=FALSE}

library(tidyverse)   # adding variables, other data manipulation
library(lubridate)

db1 <- read.table("https://raw.githubusercontent.com/AMNakamura/miscellanea/master/datasets/FakeCohort1.txt",sep="|",header=T) %>% 
  group_by(ID,YR,MO,GRP) %>%
  summarise(IND = sum(ind1)) %>%
  dplyr::select(ID,YR,MO,GRP,IND) %>%
  mutate(dt = as.Date(paste(YR,MO,"01",sep="-"))) # set to first of the month for simplicity

```

# Member Retention from Year to Year

The next step collapses the data frame on member ID and Year and creates lists of nodes and edges. 

## Data preparation

Create an aggregate dataframe to use in the creation of nodes (vertices) and edges. Use a sample for "Group 2". 

```{r, message = FALSE, warning=FALSE}

df <- subset(db1,GRP == "GRP2") %>% 
  ungroup() %>%
  dplyr::filter(IND > 0) %>%
  group_by(ID,YR) %>%
  summarize(IND=mean(IND)) %>%
  ungroup() 

```


## Add node information 

Add statistics or other information to describe the nodes (e.g., year). These can be used to define visual elements (e.g., size and color). 

```{r, message = FALSE, warning=FALSE}

node.stats <- df %>%
  group_by(YR) %>%
  summarize(member.ttl = n_distinct(ID),
            ind.mu     = mean(IND),
            ind.med    = median(IND)) %>%
  rename(V = YR)

NodeCols <- unique(node.stats$V)

last.per <- max(node.stats$V)

nodeInfo <- as.data.frame(NodeCols,stringsAsFactors = FALSE) %>%
  rename(V = NodeCols) %>%
  left_join(.,node.stats,by="V")
```


## Create the edge list 

The following performs a self-join to create an edge list, with all pairwise year combinations. Singletons are allowed, but not for the last period. Edges are restricted to sequential pairs. 


```{r, message = FALSE, warning=FALSE}

# Perform a self-join to create an edge list, with all pairwise combinations of year.
# Allow singletons for all but the last period, but remove non-sequential pairs.

e0 <- df %>% select(ID, YR,IND) %>%
  inner_join(., select(., ID,YR,IND), by = "ID") %>%
  rename(FROM = YR.x, TO = YR.y, INIT=IND.x) %>%
  group_by(ID) %>%
  mutate(rows = n()) %>%
  filter((rows == 1 & FROM == TO)| TO-FROM ==1 ) %>%
  filter(FROM != last.per) %>%
  unique %>%
  arrange(ID, FROM,TO) %>%
  dplyr::select(ID,FROM,TO,INIT)

# Aggregate

e1 <- e0 %>%
  group_by(FROM,TO) %>%
  summarize(VALUE=n()) %>% # members
  ungroup() %>%
  select(FROM,TO,VALUE) %>%
  as.data.frame()


```


# Graphs

## Network Flow

The following uses the **networkD3** and **igraph** packages to create a directed graph and simple network object. The node size reflects the number of cohort members. 

```{r flowplot, fig.cap = "Node size reflects relative cohort size", warning=FALSE, message=FALSE}


library(networkD3)
library(igraph)

g    <- graph_from_data_frame(e1,directed=TRUE,vertices = nodeInfo)
g_d3 <- igraph_to_networkD3(g)


plot(g, vertex.size = node.stats$member.ttl/5, main = "Cohort Flow from Year to Year")

```

## Sankey Diagram for Year to Year Changes

A Sankey diagram describes the flow from one set of attributes (e.g., a phase, cohort, class, development iteration) to another.

- Vertical bars show the relative size of the cohort.
- Light grey connectors show the relative size of cohort members moving from one year to another.
- Grey circles show the relative number of members who only participated for a single year.


```{r snky, fig.cap = "Node size reflects relative cohort size", fig.height=6, fig.width=10}

sankeyNetwork(g_d3$links, Nodes=g_d3$nodes,
              Source = "source", Target="target",
              Value = "value",NodeID = "name",
              fontSize=28,
              sinksRight = FALSE,
              nodeWidth = 20, nodePadding = 20,
              units ="members")


```


## Network Graphs to Visualize Movement Between Groups

```{r, message=FALSE,warning=FALSE}

df <- subset(db1) %>% 
  ungroup() %>%
  group_by(ID,GRP) %>%
  summarize(IND     = mean(IND)) %>%
  ungroup()

# Node Info (V for vertex)

node.stats <- df %>%
  group_by(GRP) %>%
  summarize(member.ttl = n_distinct(ID),
            ind.mu     = mean(IND),
            ind.med    = median(IND)) %>%
  rename(V = GRP)

NodeCols <- unique(node.stats$V)

last.per <- max(node.stats$V)

nodeInfo <- as.data.frame(NodeCols,stringsAsFactors = FALSE) %>%
  rename(V = NodeCols) %>%
  left_join(.,node.stats,by="V")

# Edge List prep

e0 <- df %>% select(ID, GRP,IND) %>%
  inner_join(., select(., ID,GRP,IND), by = "ID") %>%
  rename(FROM = GRP.x, TO = GRP.y, INIT=IND.x) %>%
  group_by(ID) %>%
  mutate(rows = n()) %>%
  filter(rows == 1 & FROM == TO|FROM != TO ) %>%
  unique %>%
  arrange(ID, FROM,TO) %>%
  dplyr::select(ID,FROM,TO,INIT)

# Aggregate

e1 <- e0 %>%
  group_by(FROM,TO) %>%
  summarize(VALUE=n()) %>% # members
  ungroup() %>%
  select(FROM,TO,VALUE) %>%
  as.data.frame()

```


- The size of the nodes describes the relative value of `ind` for the group (e.g., `ind` could reflect costs or other inputs or outputs).
- The thickness of the connecting lines reflects the number of members moving from one group to the other.

```{r network, fig.cap = "Node size reflects relative cohort size"}

g    <- graph_from_data_frame(e1,directed=TRUE,vertices = nodeInfo)

d <- as_adjacency_matrix(g) 

network <- graph_from_adjacency_matrix(d , mode='directed', diag=F )


# IBM Design Colorblind safe

clrs <- c("#648fff", "#785ef0", "#dc267f", "#fe6100" )

plot(network, layout=layout_components, 
     edge.width = E(g)$VALUE/50,
     vertex.size = nodeInfo$ind.med/20,
     vertex.color = clrs,
     edge.label = E(g)$VALUE,
     edge.curved = .5,
     main = "Flow of Members Between Groups\n Node size reflects median values for `ind`")


```


### Alluvial Charts

The following charts the flow between groups over all cohort years. 

```{r alluvial,message=FALSE, warning=FALSE}  

# Extract the data and coerce it into alluvial form.

df <- db1 %>%
  group_by(ID,YR) %>%
  summarize(GRP = last(GRP)) %>%
  ungroup() %>%
  group_by(ID) %>%
  mutate(Exit.YR = last(YR) + 1,
         Enter.Yr = first(YR) - 1) %>%
  ungroup()

df.enter <- df %>% select(ID,Enter.Yr) %>%
  mutate(GRP = "Enter") %>%
  rename(YR = Enter.Yr)

df.exit <- df %>% select(ID,Exit.YR) %>% 
  mutate(GRP = "Exit") %>%
  rename(YR = Exit.YR)


df.new <- rbind.data.frame(df[,1:3],df.enter,df.exit) 

df.plot <- df.new %>%
  mutate(YR = as.character(YR),
         YR = factor(YR, levels=c(sort(unique(df.new$YR)))),
         GRP = factor(GRP, levels = c("Enter","GRP1","GRP2","GRP3","GRP4","Exit")) ) %>%
  unique()

clrs <- c("#ffffff","#648fff", "#785ef0", "#dc267f", "#fe6100","#808080" )

library(ggalluvial)

ggplot(df.plot,
       aes(x = YR, stratum = GRP, alluvium = ID,
           fill = GRP, label = GRP,)) +
  scale_fill_manual(values=clrs) +
  geom_flow(stat = "flow", lode.guidance = "frontback",
            color = "darkgray") +
  geom_stratum() +
  ggtitle("Growth and Attrition by Year and Group Cohort") +
  theme(legend.position = "bottom",
        legend.title=element_blank()) +
  theme_minimal()

```


# Packages

Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686

Garrett Grolemund, Hadley Wickham (2011). Dates and Times Made Easy with lubridate. Journal of Statistical Software, 40(3), 1-25. URL: https://www.jstatsoft.org/v40/i03/.

Csardi G, Nepusz T: The igraph software package for complex network research, InterJournal, Complex Systems 1695. 2006. https://igraph.org

J.J. Allaire, Christopher Gandrud, Kenton Russell and CJ Yetman (2017). networkD3: D3 JavaScript Network Graphs from R. R package version 0.4. https://CRAN.R-project.org/package=networkD3

Brunson JC (2017). “ggalluvial: Layered Grammar for
Alluvial Plots.” _Journal of Open Source Software_,
*5*(49), 2017. doi: 10.21105/joss.02017 (URL:
https://doi.org/10.21105/joss.02017).


# References

