---
title: "Make Fake Survey Data 1"
author: "Ann Nakamura"
date: "orig: 7/15/2017. updates: 8/2/2021"
output: github_document
bibliography: packages.bib 
---

# Make some fake survey data

The following code makes use of [MLRun's demo-stocks "reviews"](https://github.com/mlrun/demo-stocks) dataset for sentiment analysis, topic modeling, and other analyses of text, adding some additional columns with random rating panels. 

```{r, warning=FALSE, message=FALSE}
GRPS  <- c("Business Operations","Technical and Scientific")


# Obtain some comment fields from the "reviews" list. Divide into five separate columns that can be used as dummy free responses (four L's plus one overall comment). 

# Load file
raw <- read.csv(file = 'https://raw.githubusercontent.com/mlrun/demo-stocks/master/data/reviews.csv',nrows=500)

# Split and recreate
lst  <- list()

for(i in 1:5){
  n1 <- (i-1)*100 + 1
  n2 <- (i)*100  
  print(paste(n1,n2,sep="-"))
  lst[[i]] <- raw[n1:n2,c(3)]  
}

raw.new <- do.call(cbind, lst)

# Add some additional columns, convert from matrix to data frame

suppressMessages(library(dplyr))


df <- as.data.frame(raw.new)
df$ID <- seq(1:nrow(df)) 

df <- df %>%
  mutate(grp = sample(GRPS,  nrow(df),replace=TRUE),
         YrStarted   = sample(2010:2017, nrow(df), replace=TRUE)) 

# Add eight new columns with random ratings
for(i in 1:8) {                                   
  tmp <- sample(1:8, nrow(df), replace = TRUE)    # tmp column
  df[ , ncol(df) + 1] <- tmp                    # Append to the end
  colnames(df)[ncol(df)] <- paste0("Q16_", i)     # Increment name
}

# Add another eight new columns with random ratings
for(i in 1:8) {                                   
  tmp <- sample(1:8, nrow(df), replace = TRUE)    # tmp column
  df[ , ncol(df) + 1] <- tmp                    # Append to the end
  colnames(df)[ncol(df)] <- paste0("Q18_", i)     # Increment name
}

# Introduce some missing values in selected fields. 

df[,c(1:5,8:24)] <- apply (df[,c(1:5,8:24)], 2, function(x) {x[sample( c(1:nrow(df)), floor(nrow(df)/25))] <- NA; x} )

```

# Data Checks

The following prints out quick data profiles to serve as sanity checks before using the fake data for any actual testing. 

```{r, warning=FALSE, message=FALSE, results='asis'}
suppressMessages(library(inspectdf))
suppressMessages(library(tidyverse))


inspect_types(df) %>% show_plot()
inspect_na(df) %>% show_plot()

```

# Customizing labels

Customize as needed to mimic data extracted from sources like SPSS or the Qualtrics API, where data will be labeled already. For Qualtrics, the labels will match the survey question text.   

The sample below applies labels used for a 4L's retrospective and identically labeled panels for an Importance-Satisfaction matrix. 

```{r, message=FALSE, warning=FALSE, results='asis'}
suppressMessages(library(Hmisc))

var.labels = 
  c(V1 = "I really liked", 
    V2 = "I've learned", 
    V3 = "This could have been better",
    V4 = "I would have really liked", 
    V5 = "Thinking back to your experiences in the last six months, do you have any general comments?",
    YrStarted = "What year did you join the project", 
    
    Q16_1="Component1",
    Q16_2="Component2",
    Q16_3="Component3",
    Q16_4="Component4",
    Q16_5="Component5",
    Q16_6="Component6",
    Q16_7="Component7",
    Q16_8="Component8",
    
    Q18_1="Component1",
    Q18_2="Component2",
    Q18_3="Component3",
    Q18_4="Component4",
    Q18_5="Component5",
    Q18_6="Component6",
    Q18_7="Component7",
    Q18_8="Component8")

label(df) = as.list(var.labels[match(names(df),names(var.labels))])
```

```{r,warning=FALSE,echo=FALSE, results='asis'}

suppressMessages(library(summarytools))
suppressMessages(library(webshot))


print(dfSummary(df,
                varnumbers = FALSE,
                valid.col  = FALSE,
                graph.magnif = 0.75), 
      method = "browser", file = "MkSrvy1DfSummary.html")

webshot( "MkSrvy1DfSummary.html", "MkSrvy1DfSummary.png")


```

```{r, message=FALSE,echo=FALSE}

# Credit where credit due

knitr::write_bib(c("dplyr","inspectdf","Hmisc","summarytools","tidyverse"),file="packages.bib")
```
# References

---
nocite: '@*'
---

